Scaffolder - Software for Reproducible Genome Scaffolding.

### 2011-03-08 src/article.tex
Revise background section

The advent of high throughput sequencing methods has led to a renewed interest
in algorithms to solve the problem of genome assembly \cite{pop2008,pop2009}.
Nonetheless, due to the complexity of merging large numbers of reads genome
assembly software may be unable to produce a complete sequence from these
reads. Instead the algorithm may generate several large assembled regions of
sequence (`contigs') composed of the many available individual reads that
represent the assembly of reads into the longest possible sequences, given the
data. Nonetheless, these contigs represent a fragmented picture of the genome
and require additional work to join together into a complete sequence. \pb

The process of ameliorating a draft sequence into a finished continuous
sequence can be expensive in terms of time and laboratory effort, although the
genomic data present in a set of generated contigs may be sufficient for many
research questions \cite{branscomb2002}. Nevertheless, a continuous
high-quality `finished' genome sequence does provide a greater depth of
information, such as complete resolution of repeat regions and precise
estimates of distances between genomic elements \cite{parkhill2002,fraser2002}.
The process of joining these contigs together to form a continuous genome
sequence is called the `scaffolding' or `finishing' stage. \pb

@@

Scaffolding is the process of joining a series of disconnected contigs into
a complete continuous genome sequence. Due to genomic complexity and missing
data, scaffolding may not ultimately produce a final completed sequence, but
may still succeed in a greater subset of contigs together or resolving gaps. An
overview of the required steps in the scaffolding process is outlined below:

@@

The sequencing process generates reads from either strand of the DNA helix and
the resulting contigs constructed from these reads may represent either the
5'~$\rightarrow$~3' or 3'~$\rightarrow$~5' orientation with respect to the
origin of replication. Orientating all contigs to point in the same direction
requires reverse complementing sequences where necessary. In the case of
microbial genomes this orientation will be to the 5'~$\rightarrow$~3' direction
leading away from the origin of replication.

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence, which is useful for representing both the sequenced and
to-be-resolved areas in the genome sequence.

@@

Synteny methods for scaffolding compare the assembled contigs to a more
complete reference genome sequence by searching for areas of sequence
similarity between the two. Any areas of corresponding sequence with the
reference genome can be used to infer contig placement and build the contigs
into a scaffold \cite{richter2007,zhao2008,assefa2009}. Recombination within
each of the two compared genomes can, however, reduce the effectiveness of this
analysis. \pb

Repeat regions may also be responsible for multiple gaps when building a genome
sequence; tandemly repeated nucleotide regions in the genome produces multiple
reads with similar sequence. As many assembly algorithms rely on sequence
overlaps between reads to build a contig the similarity between repeat-region
reads can result in the assembly collapsing into an artificially short sequence
at this region or being ignored by more conservative assembly algorithms. Such
regions can be resolved by using algorithms to specifically reassemble the
collapsed repeat region correctly \cite{mulyukov2002,koren2010}, while
a related approach uses unassembled sequence reads matching the regions around
a scaffold gap to construct a uniquely overlapping set of reads across it
\cite{tsai2010}. \pb

Paired-read data can provide an extra level of information about how contigs
may be scaffolded together. Heuristic scaffolding algorithms take advantage of
this data to search for optimal configuration of contigs in the scaffold using
paired-read distances \cite{dayarian2010,boetzer2011}. Synteny data from
a reference genome can also be combined with this paired-read data to estimate
the best contig configuration \cite{pop2004}. \pb

These described methods provide a wide array of approaches for merging contigs
into a larger, continuous scaffold sequence. Nevertheless the scaffolding
process may still require inserting additional sequences or further joining
contigs using PCR-derived sequence. This manual stage in the finishing
process may require moving and editing large blocks of nucleotide text by hand
which is subject to human error and precludes any reproducibility of the steps
in this process. \pb

>>>>

The advent of high throughput sequencing methods has led to a renewed interest
in algorithms to solve the problem of genome assembly \cite{pop2008,pop2009}.
Nonetheless, due to the complexity of merging large numbers of overlapping
reads genome assembly software may be unable to produce a complete sequence.
Instead the algorithm may generate several large assembled regions of sequence
(`contigs') composed of the many individual reads that represent assembly into
the longest possible sequences, given the data. Nonetheless, these contigs
represent a fragmented picture of the genome and require additional work to
join together into a complete sequence. \pb

The process of ameliorating a draft sequence into a finished continuous
sequence can be expensive in terms of time and laboratory effort, although the
genomic data present in a set of generated contigs may be sufficient for many
research questions \cite{branscomb2002}. Nevertheless, a continuous
high-quality `finished' genome sequence does provide a greater depth of
information, such as complete resolution of repeat regions and precise
estimates of distances between genomic elements \cite{parkhill2002,fraser2002}.
The process of joining these contigs together to form a continuous genome
sequence is called the `scaffolding' or `finishing' stage and is the focus of
the software described in this article. \pb

@@

Scaffolding is the process of joining a series of disconnected contigs into
a complete continuous genome sequence. Due to genomic complexity and missing
data, scaffolding may not ultimately produce a final completed sequence, but
may still succeed in joining a subset of contigs together or resolving gaps
between contigs. An overview of the required steps in the scaffolding process
is outlined below:

@@

The sequencing process generates reads from either strand of the DNA helix and
the resulting contigs constructed from these reads may represent either the
5'~$\rightarrow$~3' or 3'~$\rightarrow$~5' DNA strands with respect to the
origin of replication. Orientating all contigs to point in the same direction
requires reverse complementing sequences where necessary. In the case of
microbial genomes this orientation will be to the 5'~$\rightarrow$~3'
direction following the direction of genome replication.

@@

Given the correct order and orientation, determining the distance between
contigs results in an estimate of the complete genome size. The size of any
inter-contig gaps represents the length of an unknown region in the genome.
Filling these regions with unknown nucleotide characters (`N') allows a draft
continuous sequence, which is useful for representing both the known and
to-be-resolved areas in the genome sequence.

@@

Examples of \emph{in silico} methods include using synteny to compare the
assembled contigs to a complete reference genome sequence by searching for
areas of sequence similarity between the two. Any areas of corresponding
sequence with the reference genome can be used to infer contig placement and
build the contigs into a scaffold \cite{richter2007,zhao2008,assefa2009}.
Recombination within each of the two compared genomes can, however, reduce the
effectiveness of this analysis. \pb

Repeat regions may also be responsible for multiple gaps when building
a genome sequence; tandemly repeated nucleotide regions in the genome produce
multiple reads with similar sequence. As many assembly algorithms rely on
sequence overlaps between reads to build a contig, the similarity between
repeat-region reads can result in the assembly collapsing into an artificially
short sequence or being ignored by more conservative assembly algorithms. Such
regions can be resolved by using algorithms specifically reassemble the
collapsed repeat region correctly \cite{mulyukov2002,koren2010}, while
a related approach uses unassembled sequence reads matching the regions around
a scaffold gap to construct a uniquely overlapping set of reads across it
\cite{tsai2010}. \pb

Paired-read data can provide an extra level of information about how contigs
may be scaffolded together. Heuristic scaffolding algorithms take advantage of
this data to search for the optimal configuration of contigs in the scaffold
that matches these paired-read distances \cite{dayarian2010,boetzer2011}.
Synteny data from a reference genome can also be combined with this
paired-read data to estimate the best contig configuration \cite{pop2004}. \pb

These described \emph{in silico} methods provide a wide array of approaches
for merging contigs into a larger, continuous scaffold sequence. Nevertheless
the scaffolding process may still require manually inserting additional
sequences or further joining contigs using PCR-derived sequence. Moving and
editing large blocks of nucleotide text by hand however possibly introduces
human error and precludes any reproducibility of the steps in this process. \pb
